Pipit on the Post: Proving Pre- and Post-conditions of Reactive Systems

### 2024-01-13 2024pipit/section/01-Introduction.tex
evaluation

\begin{itemize}
  \item we motivate the need to combine manual and automated proofs of reactive systems with a strong specification language (\autoref{s:motivation});
  \item we introduce Pipit, a minimal reactive language that supports rely-guarantee contracts and properties; crucially, proof obligations are annotated with a status --- \emph{valid} or \emph{deferred} --- allowing proofs to be delayed until more is known of the program context (\autoref{s:core});
  \item we describe a \emph{checked semantics} for Pipit, which is parameterised by the property status; after checking deferred properties, programs can be \emph{blessed}, and their properties lifted to valid status (\autoref{s:core:checked});
  \item we describe an encoding of transition systems that can express under-specified rely-guarantee contracts as functions rather than relations; composing functions results in simpler transition systems (\autoref{s:transition});
  \item we identify the invariants and lemmas required to prove that the abstract transition system is an abstraction of the original semantics (\autoref{s:core:causality}, \autoref{s:transition:proof});
  \item similarly, we offer a mechanised proof that the executable transition system preserves the original semantics (\autoref{s:extraction});
  \item finally, we evaluate Pipit by implementing a time-triggered CAN bus driver, which we have partially verified (\autoref{s:evaluation}).
\end{itemize}

>>>>

\begin{itemize}
  \item we motivate the need to combine manual and automated proofs of reactive systems with a strong specification language (\autoref{s:motivation});
  \item we introduce Pipit, a minimal reactive language that supports rely-guarantee contracts and properties; crucially, proof obligations are annotated with a status --- \emph{valid} or \emph{deferred} --- allowing proofs to be delayed until more is known of the program context (\autoref{s:core});
  \item we describe a \emph{checked semantics} for Pipit, which is parameterised by the property status; after checking deferred properties, programs can be \emph{blessed}, and their properties lifted to valid status (\autoref{s:core:checked});
  \item we describe an encoding of transition systems that can express under-specified rely-guarantee contracts as functions rather than relations; composing functions results in simpler transition systems (\autoref{s:transition});
  \item we identify the invariants and lemmas required to prove that the abstract transition system is an abstraction of the original semantics (\autoref{s:core:causality}, \autoref{s:transition:proof});
  \item similarly, we offer a mechanised proof that the executable transition system preserves the original semantics (\autoref{s:extraction});
  \item finally, we evaluate Pipit by implementing the high-level logic of a time-triggered Controller Area Network (CAN) bus driver, which we have partially verified (\autoref{s:evaluation}).
\end{itemize}

### 2024-01-13 2024pipit/section/20-Motivation.tex
evaluation

To introduce Pipit, we consider a driver with a static schedule of \emph{triggers}, or actions to be performed at a particular time; this driver is a simplification of the time-triggered CAN bus specification \cite{fuehrer2001time} which we will discuss further in \autoref{s:evaluation}.
% In this system, time is partitioned into a sequence of \emph{cycles}; at the start of each new cycle, the time is reset to zero.
% Triggers also have a cycle offset and a repeat factor, which together determine the cycles in which the trigger is active.

@@

% The trigger-fetch logic requires such a combination of automatic and manual proofs.
We can now implement the trigger-fetch logic, which maintains the index that refers to the next-scheduled trigger.
The trigger-fetch logic uses the \emph{count_when} streaming function to define the index; the increment flag given to \emph{count_when} will increment the index if the previous index has expired or is inactive in the current cycle.
We simplify our presentation here and only consider a single cycle in isolation: the real system presented in \autoref{s:evaluation} has some extra complexity such as resetting the index, incrementing the cycle index at the start of a new cycle, and using machine integers.

\begin{tabbing}
  @MM@\= @MM@ \= @let index@ \= \kill
  @let@ trigger_fetch ($\textit{cycle}$: $\NN$) ($\textit{time}$: stream $\NN$): stream $\NN$ = \\
    \> $@rec @ \textit{index}.$ \\
    \> \> $@let @ \textit{inc} = \text{false} @ fby @ ((\text{time_mark}~\textit{index}) \le \textit{time} ~\vee~ \neg (\text{enabled}~\textit{index}~\textit{cycle})) @ in@$\\
    \> \> $@let @ \textit{index} = \mbox{count_when}_{\PSValid}~\text{trigger_count}~\textit{inc} @ in@$ \\
    \> \> $\text{pose}_1~(\text{lemma_can_reach_next}~\textit{cycle})~\textit{index}$; \\
    \> \> $\xcheckP{\PSUnknown}{(\text{can_reach_next_active}~\textit{cycle}~\textit{time}~\textit{index})}$; \\
    \> \> $\textit{index}$\\
\end{tabbing}

The \emph{trigger_fetch} function takes a static cycle index and the stream denoting the current time.
The increment flag and the index are mutually dependent --- the increment flag depends on the previous value of the index, while the index depends on the current value of the increment flag --- so we introduce a recursive stream for the index.
We allow the index to go one past the end of the array to denote that there are no more triggers.

>>>>

To introduce Pipit, we consider a driver with a static schedule of \emph{triggers}, or actions to be performed at a particular time; this driver is a simplification of the time-triggered Controller Area Network (CAN) bus specification \cite{fuehrer2001time} which we will discuss further in \autoref{s:evaluation}.
% In this system, time is partitioned into a sequence of \emph{cycles}; at the start of each new cycle, the time is reset to zero.
% Triggers also have a cycle offset and a repeat factor, which together determine the cycles in which the trigger is active.

@@

% The trigger-fetch logic requires such a combination of automatic and manual proofs.
We can now implement the trigger-fetch logic, which keeps track of the current trigger.
The trigger-fetch logic uses the \emph{count_when} streaming function to define the index of the current trigger; we tell \emph{count_when} to increment the index whenever the previous index has expired or is inactive in the current basic cycle.
We simplify our presentation here and only consider a single cycle in isolation: the real system presented in \autoref{s:evaluation} has some extra complexity such as resetting the index, incrementing the cycle index at the start of a new cycle, and using machine integers.

\begin{tabbing}
  @MM@\= @MM@ \= @let index@ \= \kill
  @let@ trigger_fetch ($\textit{cycle}$: $\NN$) ($\textit{time}$: stream $\NN$): stream $\NN$ = \\
    \> $@rec @ \textit{index}.$ \\
    \> \> $@let @ \textit{inc} = \text{false} @ fby @ ((\text{time_mark}~\textit{index}) \le \textit{time} ~\vee~ \neg (\text{enabled}~\textit{index}~\textit{cycle})) @ in@$\\
    \> \> $@let @ \textit{index} = \mbox{count_when}_{\PSValid}~\text{trigger_count}~\textit{inc} @ in@$ \\
    \> \> $\text{pose}_1~(\text{lemma_can_reach_next}~\textit{cycle})~\textit{index}$; \\
    \> \> $\xcheckP{\PSUnknown}{(\text{can_reach_next_active}~\textit{cycle}~\textit{time}~\textit{index})}$; \\
    \> \> $\textit{index}$
\end{tabbing}

The \emph{trigger_fetch} function takes a static cycle index and a stream denoting the current time.
The increment flag and the index are mutually dependent --- the increment flag depends on the previous value of the index, while the index depends on the current value of the increment flag --- so we introduce a recursive stream for the index.
We allow the index to go one past the end of the array to denote that there are no more triggers.

### 2024-01-13 2024pipit/section/50-Extraction.tex
some intro
evaluation

Unfortunately, our current approach is also unsuitable for generating imperative array code, as our shallowly-embedded pure transition system requires pure arrays.
In the future, we intend to address array computations and the above work duplication by generating an explicit deeply-embedded imperative program.
We currently generate a monolithic step function for each Pipit program; an embedded foreign-function interface similar to that used by Accelerate~\cite{clifton2014embedding} may also enable separate-compilation of programs.

% \item staging: similarity to \cite{gallego2021w};

>>>>

Unfortunately, our current approach is unsuitable for generating imperative array code, as our shallowly-embedded pure transition system requires pure arrays.
In the future, we intend to address array computations and the above work duplication by generating an explicit deeply-embedded imperative program.
We currently generate a monolithic step function for each Pipit program; an embedded foreign-function interface similar to that used by Accelerate~\cite{clifton2014embedding} may also enable separate-compilation of programs.

% \item staging: similarity to \cite{gallego2021w};
\CITE{Noise* code extraction} \cite{ho2022noise} hybrid embedding

### 2024-01-13 2024pipit/section/60-Evaluation.tex
bosch
evaluation

To demonstrate the feasibility of Pipit, we have implemented and verified a simple controller.
This system controls a water-flow solenoid to fill the reservoir of a coffee machine and includes multiple safeguards to reduce the risk of flooding.
The controller has two boolean inputs: the \emph{stop} switch and the \emph{low level} indicator; it returns a boolean indicating whether to engage the solenoid.
The stop switch indicates whether the reservoir's lid is open or closed; the system should never operate while the lid is open as water could spill out.
The controller should not allow water to flow for more than a minute as this may indicate a leak; if so, the controller enters a terminal error state.
Finally, to avoid switching the solenoid too often, the controller waits for ten seconds of low water level before trying to engage:

\newcommand\estop{\textit{stop}}
\newcommand\low{\textit{low}}
\newcommand\soltry{\textit{try}}
\newcommand\error{\textit{error}}
\newcommand\solen{\textit{engage}}
\begin{tabbing}
  @MM@\= @let@ engage \= @MMMM@ \= \kill
  @let@ $\mbox{reservoir}$ (\estop{} \low{}: stream $\BB$): stream $\BB$ = \\
  \> @let@ \soltry{} \> = true\_for \textit{TEN\_SECONDS} (not \estop{} $\wedge$ \low{}) @in@ \\
  \> @let@ \error{} \> = any (true\_for \textit{ONE\_MINUTE} \soltry{}) @in@ \\
  \> @let@ \solen{} \> = \soltry{} $\wedge$ not \error{} @in@ \\
  \> @check@ (\solen{} $\implies$ not \estop{}); \\
  \> @check@ (\solen{} $\implies$ \low{}); \\
  \> \solen{}
\end{tabbing}
\pagebreak

Predicate \emph{true\_for t} is true if a signal has been true for time $t$; \emph{any} is true if a signal has ever been true.
% The \emph{true\_for t} is time-bounded past-globally and \emph{any} is past-finally.
As with the previous examples, the two properties can be automatically verified.
% by \fstar{}'s proof automation.
Pipit generates real-time C code for this example\footnote{For a video of the controller in action, see \url{https://youtu.be/6IybbQFPOl8}}.

>>>>

As a preliminary evaluation of Pipit, we have implemented the high-level logic of a time-triggered Controller Area Network (CAN) bus driver~\cite{ISO11898_4}.
The CAN bus is commonly found in safety-critical automotive and industrial settings.
The time-triggered network architecture defines a static schedule of network traffic.
All nodes on the network must adhere to the same schedule, which significantly increases the reliability of periodic messages~\cite{fuehrer2001time}.

At a high level, the schedule is described by a system matrix which consists of rows of basic cycles.
Each basic cycle consists of a sequence of actions to be performed at particular time-marks.
Actions in the schedule may not be relevant to all nodes, so each node has its own local array containing the relevant triggers; trigger actions include sending and receiving application-specific messages, sending reference messages, and triggering `watch' alerts.
The trigger action for receiving an application-specific message checks that a particular message has been received since the trigger was last executed; depending on this, the driver increments or decrements a message-status-counter, which will in turn signal an error once the upper limit is reached.
Reference messages start a new basic cycle and are used to synchronise the nodes.
Watch alerts are generally placed after the expected end of the cycle and are used to signal an error if no reference message is received.

The TTCAN protocol can be implemented in two levels of increasing complexity.
In the first level, reference messages contain the index of the newly-started cycle.
In the second level, the reference messages also contain the value of a global fractional clock and whether any gaps have occurred in the global clock, which allows other nodes to calibrate their own clocks.
We implement the first level as it is more amenable to software implementation~\cite{hartwich2002integration}.

The implementation defines a streaming function that takes a stream describing the current time, the state of the hardware, and any received messages.
It returns a stream of commands to be performed, such as sending a particular reference message.
The implementation defines a pure streaming function.
To actually interact with the hardware we assume a small hardware-interop layer that reads from the hardware registers and translates the commands to hardware-register writes, but we have not yet implemented this.
We package the driver's inputs into a record for convenience:

\begin{tabbing}
  MM \= bus_status: \= \kill
  @type@ driver_input = \{ \\
    \> local_time: \> network_time_unit; \\
    \> mode_cmd: \> option mode; \\
    \> tx_status: \> tx_status; \\
    \> bus_status: \> bus_status; \\
    \> rx_ref: \> option ref_message; \\
    \> rx_app: \> option app_message_index; \\
    \}
\end{tabbing}

Here, the local-time field denotes the time-since-boot in \emph{network time-units}, which matches the bitrate of the underlying network bus.
The mode-command is an optional field which indicates requests from the application to enter configuration or execution mode.
The transmission-status describes the status of the last transmission request and may be none, success, or various error conditions.
The bus-status describes whether the bus is currently idle, busy, or in an error state.
The two receive fields denote messages received from the bus; for application-specific messages the time-triggered logic only needs the message identifier.

The driver-logic returns a stream of commands for the hardware-interop layer to perform:

\begin{tabbing}
  MM \= enable_acks: \= \kill
  @type@ commands = \{ \\
  \> enable_acks: \> bool; \\
  \> tx_ref: \>       option ref_message; \\
  \> tx_app: \> option app_message_index; \\
  \> tx_delay: \>     network_time_unit; \\
 \}
\end{tabbing}

The enable-acks field denotes whether the hardware should respond to messages from other nodes with an acknowledgement bit; in the case of a severe error acknowledgements are disabled, as the node must not write to the bus at all.
The transmit fields denote whether to send a reference message or an application-specific message.
For application-specific messages, the hardware-interop layer maintains the transmission buffers containing the actual message payload.
To meet the schedule as closely as possible, the driver predicts the next transmission and includes a transmission delay to tell the hardware exactly when to send the next message.

The implementation includes an extension of the trigger-fetch logic described in \autoref{s:motivation}, as well as state machines for tracking node synchronisation, master status and fault handling.
We generate real-time C code as described in \autoref{s:extraction}.
We evaluated the generated C code by executing with randomised inputs and measuring the worst-case-execution-time on a Raspberry Pi Pico (RP2040) microcontroller.
The runtime of the driver logic is fairly stable: over 5,000 executions, the worst-case execution time was $114\mu{}s$, while the average was $107\mu{}s$ with a standard deviation of $2.3\mu{}s$.
Earlier work on fault-tolerant TTCAN~\cite{short2007fault} describes the required slot sizes --- the minimum time between triggers --- to achieve bus utilisation at different bus rates.
For a 125Kbit/s bus, a slot size of approximately 1,500$\mu{}s$ is required to achieve utilisation above 85 per cent, which our implementation satisfies.
For the maximum CAN bus rate of 1Mbit/s, the required slot size is $184\mu{}s$; however, as our runtime does not include the time required for the hardware-interop layer, we are not confident that our implementation will satisfy full-speed CAN bus.

Our code generation can be improved in a few ways.
A common optimisation in Lustre is to fuse consecutive if-statements with the same condition~\cite{bourke2017formally}; such an optimisation seems useful here, as our treatment of optional values introduces repeated unpacking and repacking.
Some form of array fusion~\cite{robinson2017machine} may also be useful for removing redundant array operations.
Our current extraction generates a transition-system with a step function which returns a tuple of the updated state and result.
Composing these step functions together requires repeated boxing and unboxing of this tuple; we currently rely on the \fstar{} normaliser to remove this boxing.
In the future, we plan to build on the current proofs to implement a more-sophisticated encoding that introduces less overhead.

% The Bosch M-TTCAN IP module limits to 64 trigger elements \cite{bosch2019mttcan}.

We have verified a simplification of the trigger-fetch mechanism, as presented earlier (\autoref{s:motivation}).
We plan to verify the remainder of the TTCAN implementation and publish it separately.
Prior work formalising TTCAN has variously modeled the protocol itself~\cite{saha2007finite, pan2014modeling,li2018formal},
instances of the protocol~\cite{guo2020model},
and abstract models of TTCAN implementations~\cite{leen2006modeling}, but we are unaware of any prior work that has verified an \emph{executable} implementation of TTCAN.

Separately, Pipit has also been used to implement and verify a simple real-time controller for a coffee machine reservoir control system~\cite{robinson2023pipit}.
The reservoir has a float switch to sense the water level and a solenoid to allow the intake of water.
The specification includes a simple model of the water reservoir and shows that the reservoir will not exceed the maximum level.
